---
title: " Bike Ride Data Analytic Capstone Project"
author: "Philip Aiwaneose Osime"
date: "22/12/2021"
output:
  html_document: default
  tml_document: default
  pdf_document: default
  word_document: default
---
# Business Task

The business task of this project is to convert casual members into annual members.

### Key Questions
- How do casual members and annual member use the bike share differently, if at all?
- Why would a casual member buy the annual membership?
- How can cyclist use digital media to influence casual members to become annual members?

## Steps for data cleaning and visualization of the datasets

This is the capstone project for the data cleaning,data manipulation and data visualization using R notebook. 
The dataset used for this project is extracted from Motivate International Inc with the inclusion of the data source link;{https://divvy-tripdata.s3.amazonaws.com/index.html}. 
Dataset is extracted for the last 12 upload monthly versions beginning from November 2020 to October 2021.

### 1. Installing and load  library packages :
```{r}
options(repos = list(CRAN="http://cran.rstudio.com/"))
install.packages("tidyverse")
tinytex::install_tinytex()
library(tinytex)
library(tidyverse)
library(janitor)
library(lubridate)
library(ggplot2)
```
### 2. Upload cyclist trip datasets(csv files) here:

```{r}
cyclist_trip_Nov_20 <- read_csv("cyclist_trip_11_2020.csv")
cyclist_trip_Dec_20 <- read_csv("cyclist_trip_12_2020.csv")
cyclist_trip_Jan_21 <- read_csv("cyclist_trip_01_2021.csv")
cyclist_trip_Feb_21 <- read_csv("cyclist_trip_02_2021.csv")
cyclist_trip_Mar_21 <- read_csv("cyclist_trip_03_2021.csv")
cyclist_trip_Apr_21 <- read_csv("cyclist_trip_04_2021.csv")
cyclist_trip_May_21 <- read_csv("cyclist_trip_05_2021.csv")
cyclist_trip_Jun_21 <- read_csv("cyclist_trip_06_2021.csv")
cyclist_trip_Jul_21 <- read_csv("cyclist_trip_07_2021.csv")
cyclist_trip_Aug_21 <- read_csv("cyclist_trip_08_2021.csv")
cyclist_trip_Sep_21 <- read_csv("cyclist_trip_09_2021.csv")
cyclist_trip_Oct_21 <- read_csv("cyclist_trip_10_2021.csv")
```
### 3. Wrangle data and Combine into a Single File
```{r}
colnames(cyclist_trip_Nov_20)
colnames(cyclist_trip_Dec_20)
colnames(cyclist_trip_Jan_21)
colnames(cyclist_trip_Feb_21)
colnames(cyclist_trip_Mar_21)
colnames(cyclist_trip_Apr_21)
colnames(cyclist_trip_May_21)
colnames(cyclist_trip_Jun_21)
colnames(cyclist_trip_Jul_21)
colnames(cyclist_trip_Aug_21)
colnames(cyclist_trip_Sep_21)
colnames(cyclist_trip_Oct_21)
```
### 4. Inspect dataset and view  datatype for variables in each dateset.
```{r}
str(cyclist_trip_Nov_20)
str(cyclist_trip_Dec_20)
str(cyclist_trip_Jan_21)
str(cyclist_trip_Feb_21)
str(cyclist_trip_Mar_21)
str(cyclist_trip_Apr_21)
str(cyclist_trip_May_21)
str(cyclist_trip_Jun_21)
str(cyclist_trip_Jul_21)
str(cyclist_trip_Aug_21)
str(cyclist_trip_Sep_21)
str(cyclist_trip_Oct_21)
```
### 5. Change datatype for the start_station_id and end_station_id for cyclsit_trip_Nov_20 dataset.

```{r}
cyclist_trip_Nov_20 <- cyclist_trip_Nov_20 %>%
  mutate(start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
str(cyclist_trip_Nov_20)
```
### 6. Bind cyclist_trip datasets into one large file

```{r}
cyclist_project_21 <- bind_rows(cyclist_trip_Nov_20,cyclist_trip_Dec_20,cyclist_trip_Jan_21,cyclist_trip_Feb_21,cyclist_trip_Mar_21,cyclist_trip_Apr_21,cyclist_trip_May_21,cyclist_trip_Jun_21,cyclist_trip_Jul_21,cyclist_trip_Aug_21,cyclist_trip_Sep_21,cyclist_trip_Oct_21)
```
### 7. Removing the latitude and longitude columns

```{r}
cyclist_project_21 <- cyclist_project_21 %>%
  select(-c(start_lat,start_lng,end_lat,end_lng))
head(cyclist_project_21)
```
### 8. Clean Up and Add data to Prepare For Analysis
```{r}
summary(cyclist_project_21)
nrow(cyclist_project_21)
dim(cyclist_project_21)

```

### 9. Add columns that list the date, month, and year of each ride 
### 10. This will allow us to aggregrate ride data for each month, day, or year... before completing these operations we could only aggregate at the ride level.

```{r}
cyclist_project_21$date <- as.Date(cyclist_project_21$started_at)
cyclist_project_21$month <- format(as.Date(cyclist_project_21$date), "%m")
cyclist_project_21$day <- format(as.Date(cyclist_project_21$date), "%d")
cyclist_project_21$year <- format(as.Date(cyclist_project_21$date), "%Y")
cyclist_project_21$day_of_week <- format(as.Date(cyclist_project_21$date),"%A")
```

###  11. Add a "ride_length" calculation to cyclist_project_21(in seconds)
```{r}
cyclist_project_21$ride_length <- difftime(cyclist_project_21$ended_at,cyclist_project_21$started_at)
```
###  12. Convert "ride_length" from Factor to numeric so we can run calculations on the data
```{r}
is.factor(cyclist_project_21$ride_length)
cyclist_project_21$ride_length <- as.numeric(as.character(cyclist_project_21$ride_length))
is.numeric(cyclist_project_21)
head(cyclist_project_21$ride_length)
```

### 13. Remove "bad" data
-- The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divy or where  ride_length was negative 
```{r}
cyclist_project_21_v2 <- cyclist_project_21[!(cyclist_project_21$start_station_name == "HR QR" | cyclist_project_21$ride_length < 0),]
```
### 14. Conduct Descriptive Analysis
__ Descriptive analysis on the ride_length

```{r}
Average_trip_length <- mean(cyclist_project_21_v2$ride_length,na.rm = TRUE)
Median_trip_length <- median(cyclist_project_21_v2$ride_length, na.rm = TRUE)
Longest_trip_length <- max(cyclist_project_21_v2$ride_length, na.rm = TRUE)
Shortest_trip_length <- min(cyclist_project_21_v2$ride_length, na.rm = TRUE)
head(Average_trip_length)
# we can alternatively use the summary function to get the four statistics fields above
summary(cyclist_project_21_v2$ride_length)
```
### 15. Compare members and casual users
```{r}
aggregate(cyclist_project_21_v2$ride_length ~ cyclist_project_21_v2$member_casual, FUN = mean)
aggregate(cyclist_project_21_v2$ride_length ~ cyclist_project_21_v2$member_casual, FUN = median)
aggregate(cyclist_project_21_v2$ride_length ~ cyclist_project_21_v2$member_casual, FUN = max)
aggregate(cyclist_project_21_v2$ride_length ~ cyclist_project_21_v2$member_casual, FUN = min)

```

### 16. See the average ride time by each day for members vs casual users
```{r}
aggregate(cyclist_project_21_v2$ride_length~cyclist_project_21_v2$member_casual + cyclist_project_21_v2$day_of_week, FUN = mean)
```
### 17. Order the days of the week

```{r}
cyclist_project_21_v2$day_of_week <- ordered(cyclist_project_21_v2$day_of_week, levels = c("Sunday","Monday","Tuesday","Wednesday","Thursday", "Friday","Saturday"))
# rerun step 16 again:
aggregate(cyclist_project_21_v2$ride_length~cyclist_project_21_v2$member_casual + cyclist_project_21_v2$day_of_week, FUN = mean)

```
### 18. Total Number of Rides, Average Trip length. The total trip length on the weekends 

```{r}
cyclist_project_21_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual)%>%
  summarize(number_of_rides = n(), average_duration = mean(ride_length))%>%
  arrange(member_casual)%>%
  drop_na()

cyclist_project_21_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual)%>%
  filter(day_of_week == "Saturday"| day_of_week == "Sunday") %>%
  summarize(number_of_rides = n(), average_duration = mean(ride_length))%>%
  arrange(member_casual)%>%
  drop_na()

```

+ we can see casual members on the average ride for a longer time than the annual members.
+ The annual members take more rides than the casual riders in total.



### 19. Visualize the number of rides by rider type

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
readable_y_axis_unit <- function(x){
  output <- paste0(x/1000, "k")
}
cyclist_project_21_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual,weekday)%>%
  summarize(number_of_rides = n(), average_duration = mean(ride_length))%>%
  arrange(member_casual,weekday)%>%
  drop_na()%>%
  ggplot(aes(x=weekday, y= number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Number of trip rides per day per membership status ",subtitle = "Casual members vs Annual members",caption =paste0("The data between November 2020 and October 2021") , x=" The Days of the Week", y = "The Total Number of Rides", fill = "Membership type") + scale_y_continuous(labels = readable_y_axis_unit)
```

### 20. Visualization of the average duration of all trips by days of week

```{r}
add_secs <- function(x){
  output <- paste0(x, "sec")
  return(output)
}
cyclist_project_21_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday)%>%
  summarize(number_of_rides = n(), average_duration = mean(ride_length, na.rm = TRUE)) %>%
  arrange(member_casual,weekday)%>%
  drop_na() %>%
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") + labs(title ="Average duration of all trips by days of the week", subtitle = "Casual members vs Annual members",caption =paste0("The data between November 2020 and October 2021") , x=" The Days of the Week", y = "average trip duration", fill = "Membership type")  + scale_y_continuous(labels = add_secs )
```
### 21. Export Summary File For Further Analysis
creating a csv file that we can visualize in Excel, Tableau, or presentation  Software.
```{r}
counts <- aggregate(cyclist_project_21_v2$ride_length ~ cyclist_project_21_v2$member_casual + cyclist_project_21_v2$day_of_week, FUN = mean)

write.csv(counts,'avg_ride_length.csv')

```
# Data Summary 
 
## Results

+ Annual members account for 54.3% of the total rides taken between November 2020 and October 2021 while casual riders account for 45.7% of total rides.
+ Annual member riders show consistent number of rides throughout the week. 
+ This could be because most annual member riders make use of the bike share for work purposes.
+ The casual members account for 56.86% of rides on saturday and sunday. This could be an indication that most casual members use the bike share services in the weekend mostly for recreational purposes.
+ The average duration of trip for the causal rider is 34 minutes while the average duration of trip for the annual member is 14 minutes.

# Recommendations

## 1. Create an annual weekend membership. 
Giving one weekend free trial period to increase the number of casual riders signing up for an annual plan.
One free month for members who recommend riders plans. This to increase the total number of riders .

## 2. Engage the casual riders 
Produce an ad that target casual weekend riders as more casual riders make trips on weekends, giving 15% promotion bonus to casual riders that sign up for the annual membership plan.

## 3. Create a survey to investigate why casual riders are not signing up for annual plan
Create shorts surveys asking what will make casual riders sign up for an annual plan.

## 4. Target highest traffic routes for advertisement
Run an ad campaign  to target residence of high traffic congestion; enlisting the benefits of taking up an annual plan to make their life stress free and always be on time for work and appointments.




